Meta-analysis of composite outcomes in clinical trials

### 2025-01-22 MSc_poster_template.Rmd
First look, with methodologies, early results, and future direction

---
title: MSc Health Data Science Project Title
author:
  - name: Athira Mohan
  - name: Brendan Woods
    affil: 1
affiliation:
  - num: 1
    address: School of Mathematical and Statistical Sciences, University of Galway
logoleft_name: https&#58;//sport.universityofgalway.ie/assets/img/logo2.png

poster_height: "594mm"
poster_width: "841mm"
column_numbers: 4
column_margins: "0.5in"	
primary_colour:	"#a80050"	# Secondary colour use for poster design.
secondary_colour:	"#84003d"	# Main colour used for poster design.
accent_colour:	"#e6007e"	# A third colour option.
author_textcol: "white"
title_textsize: "90pt"
author_textsize: "60pt"
body_textsize: "25pt"
output: 
  posterdown::posterdown_html:
    self_contained: true
    pandoc_args: --mathjax
    number_sections: false
knit: pagedown::chrome_print
---

@@

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      tidy = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      out.width = "100%")
options(knitr.table.format = "html") 
# Load any additional libraries here
library(tidyverse)
library(plotly)
library(kableExtra)
library(posterdown)
load("\\Users\\brend\\Downloads\\data")
tmpDat$binary_insulin<-0
tmpDat$binary_insulin[tmpDat$InsulinYN == "Yes"]<-1

thesis.df<-tmpDat
```

@@

# Early Results / Descriptive Statistics of Datasets

@@

```{r morefigs, out.width='80%', echo=FALSE, fig.cap='Amazing, right?!', fig.height=4}
continuous_permutation_function<-function(x){
  permutation_t_value<-numeric(100000)
  i <-1
  repeat{
    thesis.df$Group<-sample(thesis.df$Group, 535, replace = F)
    permutation_result<-t.test(x[thesis.df$Group == "Metformin"], x[thesis.df$Group == "Placebo"])
    permutation_t_value[i]<-permutation_result$statistic
    i<- i +1
    if(i > 100000) break}
  return(permutation_t_value)
}
head(thesis.df)

@@

You can even make your plots interactive for the HTML version of the poster. You can use the HTML poster for the presentation session, and the PDF poster will be printed - so be sure the static version looks okay.

```{r interactiveplot, out.width='80%', echo=FALSE, fig.cap='Amazing, right?!', fig.height=4}

gapminder <- read_csv("https://raw.githubusercontent.com/datavizpyr/data/master/gapminder-FiveYearData.csv")
df2 <- gapminder %>%
  filter(year %in% c(1952,2007)) %>%
  filter(continent=="Asia")

df2 <- df2 %>%
  mutate(paired = rep(1:(n()/2),each=2),
         year=factor(year))

p <- df2 %>% 
  ggplot(aes(x= lifeExp, y= country)) +
  geom_line(aes(group = paired)) +
  geom_point(aes(color=year), size=4) +
  theme_bw() +
  theme(legend.position="top")

ggplotly(p)

@@

We plan to conduct further analysis using:

-   Variable discombobulation [^3]
-   Expand our minds with explosive machine learning [^4].

[^3]: Massey et al. 2005 doi: 15.36.413

[^4]: Smith et al. 1991 doi: 12.36.486

We will use the `plasticanalysis` package for this.

@@

The code and datasets for this project can be viewed at our GitHub repository here: <https://github.com/>

>>>>

---
title: Meta-analysis of composite outcomes in clinical trials 
author:
  - name: Athira Mohan
  - name: Brendan Woods
    affil: 1
affiliation:
  - num: 1
    address: School of Mathematical and Statistical Sciences, University of Galway
logoleft_name: https&#58;//sport.universityofgalway.ie/assets/img/logo2.png

poster_height: "594mm"
poster_width: "841mm"
column_numbers: 4
column_margins: "0.5in"	
primary_colour:	"#a80050"	# Secondary colour use for poster design.
secondary_colour:	"#84003d"	# Main colour used for poster design.
accent_colour:	"#e6007e"	# A third colour option.
author_textcol: "white"
title_textsize: "80pt"
author_textsize: "60pt"
body_textsize: "25pt"
output: 
  posterdown::posterdown_html:
    self_contained: true
    pandoc_args: --mathjax
    number_sections: false
knit: pagedown::chrome_print
---

@@

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      tidy = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      out.width = "100%")
options(knitr.table.format = "html") 
# Load any additional libraries here
library(tidyverse)
library(plotly)
library(kableExtra)
library(posterdown)
library(ggplot2)
load("\\Users\\brend\\Downloads\\data")
tmpDat$binary_insulin<-0
tmpDat$binary_insulin[tmpDat$InsulinYN == "Yes"]<-1

thesis.df<-tmpDat

insulin_t_value<-t.test(tmpDat$binary_insulin[tmpDat$Group == "Metformin"], tmpDat$binary_insulin[tmpDat$Group == "Placebo"])$statistic
gw32_t_value<-t.test(tmpDat$gw32[tmpDat$Group == "Metformin"], tmpDat$gw32[tmpDat$Group == "Placebo"])$statistic
gw38_t_value<-t.test(tmpDat$gw38[tmpDat$Group == "Metformin"], tmpDat$gw38[tmpDat$Group == "Placebo"])$statistic
study_squared_composite<-(insulin_t_value^2)+(gw32_t_value^2)+(gw38_t_value^2)

@@

# Descriptive Statistics of Datasets

@@

# Methodology
The meta-analysis of the EMERGE outcomes used permutation tests, which generate a null distribution by randomly shuffling the treatment group assignments of participants and their outcome variables. This breaks any true relationship between treatment and outcomes. Statistical significance is assessed by comparing the actual test statistic to the distribution of test statistics from these shuffled datasets. The P-value is calculated as the proportion of the null distribution that is more extreme than the observed test statistic.

The power of the permutation test is estimated using sampling with replacement. Participants are randomly resampled from the original data, allowing duplicates, while maintaining equal numbers from both treatment groups. The test is repeated multiple times, and power is determined by the proportion of resampled subsets yielding test statistics more extreme than the significance threshold.

# Early Results 
```{r morefigs, out.width='90%', echo=FALSE, fig.cap='Permuted Null Distribution of Composite', fig.height=6}
continuous_permutation_function<-function(x){
  permutation_t_value<-numeric(10000)
  i <-1
  repeat{
    thesis.df$Group<-sample(thesis.df$Group, 535, replace = F)
    permutation_result<-t.test(x[thesis.df$Group == "Metformin"], x[thesis.df$Group == "Placebo"])
    permutation_t_value[i]<-permutation_result$statistic
    i<- i +1
    if(i > 10000) break}
  return(permutation_t_value)
}
squared_composite<-((continuous_permutation_function(thesis.df$binary_insulin))^2) +
  ((continuous_permutation_function(thesis.df$gw32))^2) +
  ((continuous_permutation_function(thesis.df$gw38))^2)
percentile_95<-quantile(squared_composite, 0.95)

squared_composite.df<-data.frame(squared_composite)
ggplot(squared_composite.df, aes(x= squared_composite))+
  geom_histogram(bins = 200, fill = "darkslategrey", color = "darkslategrey")+
  scale_x_continuous(limits = c(0, 30)) +
  labs(x = "Squared Composite", y = "Frequency", title = "Permuted Test Statistics") +
  geom_vline(xintercept = percentile_95, color = "darkred", linetype = "dashed", size = 1)+
  annotate("text", x = percentile_95, y = 250, label = "95th Percentile", color = "darkred", angle = 90, vjust = -0.5, size = 7)+
  theme_minimal()+
  theme(axis.title = element_text( size = 24),
        axis.text = element_text(size = 18),
        plot.title = element_text(size = 28))+
  geom_vline( xintercept = study_squared_composite, color = "gold3", size = 1 )+
  annotate("text", x = study_squared_composite, y = 250, label = "EMERGE Composite", color = "gold3", angle = 90, vjust = -0.5, size = 7)

@@

```{r morefigs2, out.width='90%', echo=FALSE, fig.cap='Power of Permutation Test', fig.height=5}
APPLE<-7.881899
power_n_30<-0.219598
power_n_40<-0.264418
power_n_50<-0.314862
power_n_60<-0.367066
power_n_70<-0.418002
power_n_80<-0.468156
power_results<-c(power_n_30, power_n_40, power_n_50, power_n_60, power_n_70, power_n_80)
sample_size<-c(30, 40, 50, 60, 70, 80)

power_graph.df<-data.frame(sample_size, power_results)
ggplot(power_graph.df, aes(x=sample_size, y=power_results)) +
  geom_line( color="mediumseagreen", size=1.5, alpha=0.9, linetype=1) +
  theme_bw() + labs(title = "Power vs. Sample Size", x = "Sample Size", y= "Power")+
  scale_x_continuous(limits = c(30, NA), expand = c(0, 0))  +
  ylim(0, NA)+
  theme(axis.title = element_text(size = 24),
         axis.text = element_text(size = 18),
        plot.title = element_text(size = 28))

@@

The next steps in our project is to compare the permutation method with other
approaches to improving the analysis of composite outputs, in particular to the
methods described by Mc Menamin *et al,* [^5].

[^5]: Mc Menamin *et al,* 2018 doi:https://doi.org/10.1186/s13023-018-0819-1

@@

The code and datasets for this project can be viewed at our GitHub repository here: <https://github.com/Brendan-G-Woods/Thesis>
